---
title: "Rolling aggregate"
author: "Wei Luo"
date: "10 May 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
```

## Window functions again: RcppRoll
From Week 6, the following types window operations were mentioned:

1. Ordinal functions: `row_number`; `min_rank` and `dense_rank`; `percent_rank` and `cume_dist`; `ntile`; `first`, `nth`, and `last`.
2. Offset functions: `lead` and `lag`.
3. Cumulative aggregates: `cumsum`, `cummin`, `cummax`, `cumall`, `cumany`, and `cummean`. (See also the `accumulate` function in package `pyrrr`.)
4. Rolling aggregates

This week, we will look at the rolling aggregate functions in the package `RcppRoll`.


### How to measure temperature change?
We have seen that the air temperature has daily and yearly cycles. 
```{r}
library(tidyverse)
library(lubridate)
library(nycflights13)
weather_lga <- weather %>%
  mutate(dt =  make_date(year, month, day)) %>% 
  filter(origin == "LGA")

weather_lga %>%
  ggplot(aes(x = dt, y = temp, colour = factor(hour))) + 
  geom_line()

weather_lga %>%
  group_by(month, hour) %>% 
  summarise(t_avg = mean(temp)) %>% 
  ggplot(aes(x = hour, y = t_avg, colour = factor(month))) + 
  geom_line()
```


In [this paper](https://www.ncbi.nlm.nih.gov/pubmed/12060315), sudden change of temperature is measured by the difference between a day's temperature and the mean temperature of the previous 3 days.
 
```{r}
library(RcppRoll)
weather_lga <- weather_lga %>% 
  filter(hour == 7) %>% 
  mutate(t_3d_before = roll_meanr(lag(temp), 3)) %>% 
  mutate(tdiff_3 = temp - t_3d_before) 

weather_lga %>% 
  ggplot() +
  geom_line(aes(dt, tdiff_3))
```



> Can you modify the code to measure the difference with the *median* temperature of the previous 3 days?

```{r echo = FALSE}
weather_lga <- weather_lga %>% 
  mutate(tdiff_3_median = temp - roll_medianr(lag(temp), 3))

weather_lga %>% 
  ggplot() +
  geom_line(aes(dt, tdiff_3_median))
```

## Temperature and flight numbers
Suppose we are interested in the impact of sudden weather change on the number of daily flights. We can add that indicator into the model we developed last week.
```{r}
library(mgcv)
daily <- flights %>% 
  mutate(date = make_date(year, month, day)) %>% 
  group_by(date) %>% 
  summarise(n = n()) %>% 
  mutate(wkday = factor(wday(date, label = TRUE), ordered = FALSE ),
         date_n = as.numeric(date))


daily <- daily %>% left_join(select(weather_lga, date = dt, tdiff_3, temp))

model_gam3 <- gam(n ~ s(date_n, by = wkday) + wkday, data = daily) 
model_gam5 <- gam(n ~ s(date_n, by = wkday) + wkday + s(tdiff_3), data = daily) 
```

> Does the new model fit the data better?

```{r echo=FALSE}
AIC(model_gam3,
    model_gam5)
gam.check(model_gam5)
```


Let's have a look at the relation.
```{r}
plot(model_gam5)
```

What can you say about the effect of temperature drop?



> Can you modify the model using the difference of median temperature?

```{r echo=FALSE}
daily <- daily %>% left_join(select(weather_lga, date = dt, tdiff_3_median))
model_gam7 <- gam(n ~ s(date_n, by = wkday) + wkday + s(tdiff_3_median), data = daily) 

plot(model_gam7)

AIC(model_gam3,
    model_gam5,
    model_gam7)
gam.check(model_gam7)
```


